---
author: "Heather Parry"
title: "In class Oct 20"
output: 
  html_document:
    toc: TRUE
    toc_float: TRUE
    
  
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Here is a demonstration of how to use


## making a new code chunk

the shortcut key on mac command- option-i

If I press this key combo, RStudio will insert a new code chunk!
If I want, I can give the chunk a name

```{r how to make a new chunk}

```

## Reading a csv file

I'm going to read some data from a csv (comma-separated-values) file.

I can use read.csv() along with the path to the file to read the data into a data.frame.

Here's an example:

```{r read_birds_data}
library(here)
require(here)
dat_birds = read.csv(here("data", "hab.sta.csv"))


```

## Previewing the contents of a data.frame

I can use the head() function to print the first few lines of a data.frame.

Here's an example with the birds data

```{r head_function_demo}
head(dat_birds)



```

## Demo plot

I can use hist() to create a histogram

When I plot 
```{r simple plot}

hist(dat_birds$aspect)



```

## How to customize a histogram

I can title the argument main -


```{r customize histogram 1}
hist(dat_birds$aspect, 
     main = "Histogram of study site aspect",
     breaks = 36)

```

